### 2016-02-15 doc/thesis/intro.tex

The branches of the tree may have associated lengths, representing either
evolutionary distance or calendar time between ancestors and their descendants.
The term ``evolutionary distance'' is used here imprecisely to mean any sort of
quantitative measure of evolution, such as the number of differences between
the DNA sequences of an ancestor its descendant, or the difference in average
body mass or height. Particular examples of evolutionary distance, based on
genetic data, will be discussed below in subsection~\ref{subsubsec:genetree}. A
phylogeny with branch lengths in calendar time units is often referred to as
\defn{time-scaled}. In a time-scaled phylogeny, the internal nodes can be
mapped onto a timeline by using the tips of the tree, which usually map to the
present day, as a reference point~\autocite{nee1992tempo}. The corresponding
points on the timeline are called \defn{branching times}, and the rate of their
accumulation is referred to as the \defn{branching rate}. Rooted trees whose
tips are all the same distance from the root are called \defn{ultrametric}
trees~\autocite{buneman1974note}. These concepts are illustrated in
Figure~\ref{fig:speciestree}.

@@

\subsection{Species trees and transmission trees}
\label{subsubsec:speciestree}

A \defn{species tree} is a particular type of phylogeny in which the taxa are
species, and the branching times correspond to historical speciation events. By
speciation events, we mean times when two formerly interbreeding groups of
organisms ceased to reproduce with each other. In general, we do not have
access to the true tree relating a particular group of extant species, as this
would usually imply certain knowledge of speciation events in the distant past.
Therefore, all species trees considered by researchers are estimates, based on
the genetic or phenotypic similarity of the extant taxa, the fossil record, or
both. We discuss some of the challenges associated with this estimation in
subsection \ref{subsubsec:treeconv} below. However, for the moment, we ignore
these complications and consider what can be said about the true species tree.

Speciation often occurs by an \defn{allopatric} process, where a sub-population
of organisms is isolated from the original population by a geographic
barrier~\autocite{coyne2004speciation}. Over time, differing selection
pressures on either side of the barrier, coupled with genetic drift, cause the
two populations to diverge genetically. This eventually results in two distinct
species, which will appear as two subtrees or \defn{clades} in their species
tree if descendants of both survive to the present. Of course, this is not
necessarily the case, as selection may cause one or both of the new
sub-populations or their descendants to become extinct. On the other hand,
selection may favour one of the sub-populations, allowing it to expand into new
regions or niches and possibly paving the way for subsequent speciation events.
Speciation is not necessarily allopatric: \defn{sympatric} speciation refers to
the emergence of a new species without any geographic
barrier~\autocite{coyne2004speciation}. Indeed, there is a whole continuum of
possible speciation processes~\autocite{fitzpatrick2008if}, ranging from
completely allopatric to completely sympatric. However, at least for eukaryotic
organisms, allopatric speciation is thought to be by far the most common
type~\autocite{coyne2004speciation}.

A different type of tree used in the study of pathogens is a \defn{transmission
tree}. In such trees, tips represent infected hosts, while internal nodes
correspond to transmissions from one host to another. Transmission trees
generally have branch lengths in units of calendar time, with a root
corresponding to the index case, and branching times indicating times of
transmission. The internal nodes may be labelled with the donor of the
transmission pair, if this is known. The tips of the tree, rather than being
fixed at the present day, are placed at the time at which the pathogen
population of the individual was sampled. Consequently, the transmission tree
may not be ultrametric, but may have tips located at varying distances from the
root. Such trees are said to have \defn{heterochronous} taxa or
samples~\autocite{drummond2003measurably}, in contrast to the
\defn{isochronous} samples found in macro-organisms species trees. A
transmission tree is illustrated in Figure~\ref{fig:contactnet} (A).

At first glance, transmission trees and species trees seem to have nothing to
do with each other. One represents the evolutionary history of a group of
organisms, the other traces the spread of a disease through a host population.
However, when considering epidemics of RNA viruses, transmission trees and
species trees are highly related. The viral population within a single host has
often been referred to as a \defn{quasispecies}~\autocite{domingo2012viral},
and due to the fast evolutionary rate of these viruses, two hosts' viral
populations are invariably genetically distinct. The transmission process is
functionally identical to allopatric speciation, involving geographic isolation
of a sub-population in a new host and subsequent genetic divergence. However,
for the most part, it is not selection but rather host epidemiology which
causes certain lineages to proliferate while driving others to extinction. A
host who recovers, dies, or is isolated equates to an extinction in the
transmission tree. On the other hand, dense populations in frequent contact
often experience rapid epidemic growth, which causes similarly rapid
accumulation of branching points in the tree.

The fact that these evolutionary (ie. speciation) and epidemiological processes
occur on the same time scale for RNA viruses~\autocite{drummond2003measurably}
is the basis for the field of
\defn{phylodynamics}~\autocite{grenfell2004unifying}. The similarity between
transmission and speciation means that it is possible to apply many of the
tools and techniques developed for species trees to viral transmission trees,
and interpret the results in the context of transmission rates and host
epidemiology. As a basic example, the lineages-through-time
plot~\autocite{nee1992tempo}, which plots the historical speciation rate
against time, can be used to quantify the incidence of new infections over the
course of an epidemic~\autocite{holmes1995revealing}. In
subsection~\ref{subsubsec:treeshape} below, we review some of these methods in
more detail, as well as several which have been developed specifically for
viral phylodynamics, and discuss what these methods can tell us about host
epidemiology.

Finally, we must note two important points. First, we have not yet considered
the selection pressures operating within a single host, such as the immune
response. Second, we reiterate that both transmission trees and species trees
are theoretical objects which are almost never known with certainty for real
data. Indeed, the estimation of transmission trees is often problematic, and it
is in such estimation that within-host forces play a role. Both of these issues
are discussed further below in subsection~\ref{subsubsec:treeconv}, though we
continue to ignore them in the next subsection.

\subsection{Gene trees and viral phylogenies}
\label{subsubsec:genetree}

A \defn{gene tree} is another type of phylogeny which traces the evolutionary
history of one or more sections of genetic material. Gene trees can be built
from small parts of genes up to whole genomes, but for ease of exposition we
will use only the word ``gene''. The tips of a gene tree represent the
different \defn{alleles}, or versions, of the gene found in the studied taxa.
The branching points in a gene tree represent genetic divergence events - that
is, the times when two distinct alleles first came into existence through
mutation.

Gene trees are usually estimated using the nucleotide or amino acid sequences
of the genes in question. Although there are methods for estimating gene trees
using other types of molecular data, such as gene orders, we will focus on
sequence-based methods here. As with species trees, it is not possible to know
gene trees with certainty, only to estimate them. However, the problem of
estimating gene trees directly from data is somewhat better defined than that
of estimating species trees, because of the quantitative nature of the
molecular data used to label the tips. In general, methods to estimate gene
trees fall into three major categories: distance-based, maximum parsimony, and
model-based~\autocite{nei2000molecular}. The first two categories are generally
no longer used for phylogenetic inference, since model-based methods have been
demonstrated to be almost universally more accurate. Therefore, we will review
in detail only the last category.

Model-based methods include both maximum likelihood and Bayesian approaches. To
use them, one must first define a parametric model of sequence evolution, here
denoted $M$. For all possible trees $T$ (with branch lengths) and choices of
model parameters $\theta$, the model must define a probability density function
on observed sequence data $D$, denoted $f(D \mid T,\,\theta)$. This is also
sometimes written $\Pr(D \mid T,\,\theta)$, even though it is a probability
\emph{density}, not a probability. Maximum likelihood aims to find a particular
tree and values for the model parameters which optimize the \defn{likelihood
function} $\L$ for the observed data,
\[
  \L(\theta,\, T \mid D) = f(D \mid T,\, \theta).
\]
Bayesian methods, rather than trying to find point estimates for $\theta$ and
$T$, aim to approximate the entire \defn{posterior distribution} for the
observed data,
\[
  f(T,\,\theta \mid D) = \frac{f(D \mid T,\,\theta)f(T,\,\theta)}{f(D)}.
\]
Here, $f(T, \theta)$ denotes the prior distribution on $T$ and $\theta$, and
$f(D)$ is the marginal density of the data (that is, the integral of the
numerator over all values of $T$ and $\theta$).

A \defn{viral phylogeny} is a particular kind of gene tree which relates the
genetic sequences of viruses carried within one or more hosts. In this work, we
consider viral phylogenies constructed from one sequence per infected
individual, so that the tips of the are associated with both the viral sequence
and the corresponding host.

\subsection{Estimating transmission trees}
\label{subsubsec:treeconv}

Most applications of viral phylodynamics, including this project, aim to answer
epidemiological questions~\autocite{pybus2009evolutionary, volz2013viral}. For
example, phylodynamic methods have been used to investigate the degree of
clustering~\autocite{hughes2009molecular} and the effect of elevated
transmission risk in acute infection~\autocite{volz2012simple} (a more thorough
review is given below in subsection~\ref{subsubsec:appphylo}). For these
applications, we are ultimately interested in the transmission tree, which is a
representation of an epidemiological process (transmission), as opposed to the
viral phylogeny, which represents a biological process (viral evolution). As
discussed above in subsection~\ref{subsubsec:speciestree}, transmission trees
and are in practice unknown, and must be estimated from available data. Here,
we review some of the ways to perform such estimation.

In principle, transmission trees can be reconstructed by on-the-ground
epidemiological methods, that is, by asking each person who they contacted.
This is particularly relevant for sexually transmitted diseases such as HIV,
where individuals are more likely to recall who they contacted and
when~\autocite{klovdahl1985social}. This kind of data is challenging to collect
in detail. Most commonly, when applying phylodynamics in practice, the
transmission tree is estimated from the viral
phylogeny~\autocite{volz2013viral}. As discussed above in
subsection~\ref{subsubsec:genetree}, the viral phylogeny is itself also estimated;
however, for the moment we ignore this complication and describe methods for
estimating the transmission tree if the viral phylogeny were known.

When we consider the fact that the viral phylogeny is not known with certainty,
the methods just described follow a two-step procedure. The first step is to
estimate the viral phylogeny, and the second step is to use it to estimate the
transmission tree.

\autocite{didelot2014bayesian} develop a Bayesian version of the two-step
approach. They first estimate a time-scaled phylogeny from viral samples, then
estimate the transmissions which took place along the phylogeny. Due to
within-host evolution, the transmissions are allowed to occur anywhere along
the branches, rather than being constrained to happen at branching points.
However, the authors admit that their method requires sampling of every
infected individual, although they indicate that it could be extended to relax
this assumption.

A different approach is undertaken by Jombart
\etal~\autocite{jombart2011reconstructing}, who build transmission trees
directly from sequence data without the intermediate step of constructing a
phylogeny. Their method finds the most likely geneology among the sampled
sequences, assuming that the that the common ancestors to sampled isolates are
themselves sampled. Although this may be a realistic assumption in the early
stages of an epidemic of a slower evolving pathogen, it is unlikely for
ancestors to be present among sequences sampled from an ongoing HIV epidemic
\comment{Needs a reference}.

Several alternative approaches have been developed which aim to
incorporate available epidemiological data as well. Cottam
\etal~\autocite{cottam2008integrating} first built a viral phylogeny, then
enumerated all transmission trees consistent with this phylogeny by labelling
the internal nodes of the phylogeny in every possible way. Each of the putative
transmission trees was assigned a likelihood based on infection time and
geographic separation.

>>>>

The branches of the tree may have associated lengths, representing either
evolutionary distance or calendar time between ancestors and their descendants.
The term ``evolutionary distance'' is used here imprecisely to mean any sort of
quantitative measure of evolution, such as the number of differences between
the DNA sequences of an ancestor its descendant, or the difference in average
body mass or height. A phylogeny with branch lengths in calendar time units is
often referred to as \defn{time-scaled}. In a time-scaled phylogeny, the
internal nodes can be mapped onto a timeline by using the tips of the tree,
which usually map to the present day, as a reference
point~\autocite{nee1992tempo}. The corresponding points on the timeline are
called \defn{branching times}, and the rate of their accumulation is referred
to as the \defn{branching rate}. Rooted trees whose tips are all the same
distance from the root are called \defn{ultrametric}
trees~\autocite{buneman1974note}. These concepts are illustrated in
Figure~\ref{fig:speciestree}.

@@

\subsection{Transmission trees and epidemiology}

In epidemiology, a \defn{transmission tree} is a graphical representation of an
epidemic's progress through a population. Like phylogenies, transmission trees
have tips, nodes, edges, and branch lengths. However, rather than recording an
evolutionary process (speciation), they record an epidemiolgical process
(transmission). The tips of a transmission tree represent infected hosts, while
internal nodes correspond to transmissions from one host to another.
Transmission trees generally have branch lengths in units of calendar time,
with a root corresponding to the index case, and branching times indicating
times of transmission. The internal nodes may be labelled with
the donor of the transmission pair, if this is known. The tips of the tree,
rather than being fixed at the present day, are placed at the time at which the
individual was removed from the epidemic, such as by death, recovery,
isolation, behaviour change, or migration. Consequently, the transmission tree
may not be ultrametric, but may have tips located at varying distances from the
root. Such trees are said to have \defn{heterochronous}
taxa~\autocite{drummond2003measurably}, in contrast to the
\defn{isochronous} taxa found in the phylogenies of macro-organisms.  A
transmission tree is illustrated in Figure~\ref{fig:virustrees}.

\begin{figure}
    \centering
    \includegraphics{virustrees}
    \label{fig:virustrees}
    \caption[Illustration of a transmission tree]{
        A transmission tree representing an epidemic's progress through
        four individuals. The index case, patient $a$, transmitted the
        virus to patients $b$ and $c$; patient $c$ went on to infect
        patient $d$. 
    }
\end{figure}

Since transmission trees are essentially a detailed record of an epidemic's
progress, they contain substantial epidemiological information. As a basic
example, the \gls{ltt} plot~\autocite{nee1992tempo}, which plots the number of
lineages in a phylogeny against time, can be used to quantify the incidence of
new infections over the course of an epidemic~\autocite{holmes1995revealing}.
Many more diverse epidemiological parameters have been investigated using
transmission trees, such as the degree of
clustering~\autocite{hughes2009molecular} and the effect of elevated
transmission risk in acute infection~\autocite{volz2012simple}. However, in all
but the most well-studied of epidemics, this is not possible to obtain through
traditional epidemiological methods. The time and effort to conduct detailed
interviews and contact tracing of a sufficient number of infected individuals
is usually prohibitive. Even when the resources for such methods are available,
patients may not always recall whom they contacted and when, especially in the
case of airborne transmission. Consequently, the transmission tree must be
estimated using other methods. Most commonly, this is done by exploiting the
relationship between transmission trees and viral phylogenies.

\subsection{Estimating transmission trees from viral phylogenies}

In general, \defn{viral phylogenies} are simply phylogenetic trees relating
virus strains. \comment{Is ``strain'' the best word here?} In phylodynamics, we
often consider \defn{inter-host} phylogenies, which relate one viral genotype
from each host in a population. The tips of the tree are labelled with both the
virus and its host. The crux of
\defn{phylodynamics}~\autocite{grenfell2004unifying} is the fact that the
processes generating these two types of tree - epidemiological processes, for
transmission trees, and evolutionary processes, for viral phylogenies - occur
on similar time scales for RNA viruses~\autocite{drummond2003measurably}. As a
result, there is a close relationship between the two trees. In particular, the
transmission process is quite similar to \defn{allopatric
speciation}~\autocite{coyne2004speciation}, where genetic divergence follows
the geographic isolation of a sub-population of organisms. Thus, transmission,
which is represented as branching in the transmission tree, causes branching in
the viral phylogeny as well. Similarly, the removal of an individual causes the
extinction of their viral lineage. Due to these relationships, the topology of
the viral phylogeny is often used as a proxy for the topology of the
transmission tree. However, there are several complications and caveats which
must be kept in mind when estimating the transmission tree in this manner.

First are the issues of rooting and time-scaling. Modern likelihood-based
methods of phylogenetic reconstruction produce unrooted trees whose branch
lengths measure genetic distance in units of expected substitutions per site,
whereas transmission trees are rooted with branches measuring calendar time.
We generally assume that sampling a virus from the individual also corresponds
to their removal from the transmission tree\comment{needs a reference}, so the
positions of the tips in time are fixed. Therefore, to transform a viral
phylogeny into an estimated transmission tree, we must find a root and an
assignment of branch lengths such that the tips are placed at their proscribed
times. Of course, there are an infinite number of possible combinations of root
and branch lengths which result in the correct placement of the tips, so we
would also like to find values which are the most ``reasonable'' given our
data. By reasonable, we mean that the variation in \defn{evolutionary rate}, 
which is the ratio of a branch's length in genetic distance to its length in
calendar time, is low across the tree. While there is some variation among
hosts, due to immunological and other factors, we generally expect to observe
globally similar evolutionary rates. Methods for time-scaling a phylogeny
include root-to-tip regression~\autocite{drummond2003inference}, which we apply
in this work, and least-square dating~\autocite{to2015fast}. Both of these
methods can be used to root the tree, by simply trying all possible root
positions and choosing the one which minimizes a loss function
($1-R^2$ or root-mean-square error for root-to-tip regression, sum of squared
errors for least-square dating). Alternatively, the tree may be rooted with an
outgroup before time-scaling.

A second, perhaps more insidious problem is the fact that the correspondence
between the topologies of the viral phylogeny and transmission tree is not
necessarily exact. Due to intra-host diversity, the viral strain which is
transmitted may have split from another lineage within the donor long before
the transmission event occurred. Hence, the branching point in the viral
phylogeny may be much earlier than that in the transmission tree. Another
possibility is that one host transmitted to two or more recipients, but the
lineages they each received originated within the donor host in a different
order than that in which the transmissions occurred. In this case, the topology
of the transmission tree and the viral phylogeny will be mismatched. Although
phylodynamics is quite new, these phenomena have been studied in evolutionary
biology for some time. Viral phylogenies are a specific version of a more
general class of trees called \defn{gene trees}, which represent the
evolutionary history of a section of genetic material. Transmission trees, on
the other hand, are highly analogous to \defn{species trees}, whose tips are
species and internal nodes are common ancestors. This analogy derives from the
functional similarity between transmission and allopatric speciation. Hence,
the potential discordance between transmission trees and viral phylogenies is
the similar to that between gene and species trees, which is called
\defn{incomplete lineage sorting}. In practice, this discordance has not proven
an insurmountable problem: for example, \textcite{leitner1996accurate} were
able to accurately recover a known transmission tree using a viral phylogeny.

A final caveat is that the viral phylogeny itself is not known with certainty,
so it must also be estimated from genetic data. Phylogenetic inference is a
complex topic which we will not discuss in detail here (see e.g.
\autocite{nei2000molecular} for a full review). Most modern analyses use
model-based methods, which simultaneously estimate the phylogeny with branch
lengths and the parameters of a model of evolution. Although they usually work
well in practice \comment{needs a reference}, the estimated topology can vary
based on the model used and, in the case of Bayesian analysis, the priors.  In
addition, intra-host viral populations are genetically heterogeneous, so
choosing a single representative genotype per host is necessarily imprecise.
One can use either the genotype of a specific virion sampled from the host, or
a synthetic genotype, such as a consensus or reconstructed ancestral sequence.

What we have just discussed is a two-step procedure for estimating the
transmission tree. First, a viral phylogeny is constructed from genetic
sequence data, and then it is rooted and time-scaled into a transmission tree.
This approach is straightforward, frequently used, and has the advantage of
leveraging tried-and-true tools for phylogenetic inference. However, it also
has drawbacks, perhaps the most obvious being the multiplication of errors
produced by the separate steps. One commonly used alternative method is to
directly estimate a time-scaled phylogeny by simultaneously inferring the tree
topology, its root and branch lengths, and the parameters of a \defn{molecular
clock} model. A molecular clock is a hypothesis about the evolutionary rates
along the branches of the tree, such as that they are all equal (a \defn{strict
clock}) or that they are \gls{iid} from a common distribution (a \defn{relaxed
clock}). This inference is usually done in a Bayesian framework using
\gls{MCMC}, so that prior information (including the tip dates) \comment{are
tip dates part of the prior?} can be included in the analysis, and the
so-called nuisance parameters of the molecular clock model can be marginalized
out. Software packages for performing these analyses include
BEAST~\autocite{bouckaert2014beast} and MrBayes~\autocite{ronquist2012mrbayes}.

Several other authors have developed methods tailor-made for inferring
transmission trees. \textcite{didelot2014bayesian} develop a Bayesian version
of the two-step approach which allows transmissions to occur anywhere along the
branches of a transmission tree, rather than being constrained to the branching
points in the viral phylogeny. The method requires sampling of every infected
individual, although the authors indicate that it could be extended to relax
this assumption. \textcite{cottam2008integrating} describe a likelihood-based
method which enumerates all transmission trees consistent with an established
phylogeny, assigning each a likelihood based on other epidemiological data. 
This approach is novel in its integration of data from multiple sources,
however because it enumerates a large portion of the tree space, it is unlikely
to scale to larger epidemics. A different approach is undertaken by
\textcite{jombart2011reconstructing}, who describe a method to build
transmission trees directly from sequence data, contingent on the common
ancestors also being sampled. This makes the method attractive for
slow-evolving pathogens, but less practical for viral outbreaks where samples
from common ancestors are unlikely to be available.

%\subsection{Species trees and transmission trees}
%\label{subsubsec:speciestree}
%
%A \defn{species tree} is a particular type of phylogeny in which the taxa are
%species, and the branching times correspond to historical speciation events. By
%speciation events, we mean times when two formerly interbreeding groups of
%organisms ceased to reproduce with each other. In general, we do not have
%access to the true tree relating a particular group of extant species, as this
%would usually imply certain knowledge of speciation events in the distant past.
%Therefore, all species trees considered by researchers are estimates, based on
%the genetic or phenotypic similarity of the extant taxa, the fossil record, or
%both. We discuss some of the challenges associated with this estimation in
%subsection \ref{subsubsec:treeconv} below. However, for the moment, we ignore
%these complications and consider what can be said about the true species tree.
%
%Speciation often occurs by an \defn{allopatric} process, where a sub-population
%of organisms is isolated from the original population by a geographic
%barrier~\autocite{coyne2004speciation}. Over time, differing selection
%pressures on either side of the barrier, coupled with genetic drift, cause the
%two populations to diverge genetically. This eventually results in two distinct
%species, which will appear as two subtrees or \defn{clades} in their species
%tree if descendants of both survive to the present. Of course, this is not
%necessarily the case, as selection may cause one or both of the new
%sub-populations or their descendants to become extinct. On the other hand,
%selection may favour one of the sub-populations, allowing it to expand into new
%regions or niches and possibly paving the way for subsequent speciation events.
%Speciation is not necessarily allopatric: \defn{sympatric} speciation refers to
%the emergence of a new species without any geographic
%barrier~\autocite{coyne2004speciation}. Indeed, there is a whole continuum of
%possible speciation processes~\autocite{fitzpatrick2008if}, ranging from
%completely allopatric to completely sympatric. However, at least for eukaryotic
%organisms, allopatric speciation is thought to be by far the most common
%type~\autocite{coyne2004speciation}. 
%
%At first glance, transmission trees and species trees seem to have nothing to
%do with each other. One represents the evolutionary history of a group of
%organisms, the other traces the spread of a disease through a host population.
%However, when considering epidemics of RNA viruses, transmission trees and
%species trees are highly related. The viral population within a single host has
%often been referred to as a \defn{quasispecies}~\autocite{domingo2012viral},
%and due to the fast evolutionary rate of these viruses, two hosts' viral
%populations are invariably genetically distinct. The transmission process is
%functionally identical to allopatric speciation, involving geographic isolation
%of a sub-population in a new host and subsequent genetic divergence. However,
%for the most part, it is not selection but rather host epidemiology which
%causes certain lineages to proliferate while driving others to extinction. A
%host who recovers, dies, or is isolated equates to an extinction in the
%transmission tree. On the other hand, dense populations in frequent contact
%often experience rapid epidemic growth, which causes similarly rapid
%accumulation of branching points in the tree. 
%
%The fact that these evolutionary (ie. speciation) and epidemiological processes
%occur on the same time scale for RNA viruses~\autocite{drummond2003measurably}
%is the basis for the field of
%\defn{phylodynamics}~\autocite{grenfell2004unifying}. The similarity between
%transmission and speciation means that it is possible to apply many of the
%tools and techniques developed for species trees to viral transmission trees,
%and interpret the results in the context of transmission rates and host
%epidemiology. 
%
%%As a basic example, the lineages-through-time
%%plot~\autocite{nee1992tempo}, which plots the historical speciation rate
%%against time, can be used to quantify the incidence of new infections over the
%%course of an epidemic~\autocite{holmes1995revealing}. 
%In
%subsection~\ref{subsubsec:treeshape} below, we review some of these methods in
%more detail, as well as several which have been developed specifically for
%viral phylodynamics, and discuss what these methods can tell us about host
%epidemiology.
%
%Finally, we must note two important points. First, we have not yet considered
%the selection pressures operating within a single host, such as the immune
%response. Second, we reiterate that both transmission trees and species trees
%are theoretical objects which are almost never known with certainty for real
%data. Indeed, the estimation of transmission trees is often problematic, and it
%is in such estimation that within-host forces play a role. Both of these issues
%are discussed further below in subsection~\ref{subsubsec:treeconv}, though we
%continue to ignore them in the next subsection.
%
%\subsection{Gene trees and viral phylogenies}
%\label{subsubsec:genetree}
%
%A \defn{gene tree} is another type of phylogeny which traces the evolutionary
%history of one or more sections of genetic material. Gene trees can be built
%from small parts of genes up to whole genomes, but for ease of exposition we
%will use only the word ``gene''. The tips of a gene tree represent the
%different \defn{alleles}, or versions, of the gene found in the studied taxa.
%The branching points in a gene tree represent genetic divergence events - that
%is, the times when two distinct alleles first came into existence through
%mutation.
%
%Gene trees are usually estimated using the nucleotide or amino acid sequences
%of the genes in question. Although there are methods for estimating gene trees
%using other types of molecular data, such as gene orders, we will focus on
%sequence-based methods here. As with species trees, it is not possible to know
%gene trees with certainty, only to estimate them. However, the problem of
%estimating gene trees directly from data is somewhat better defined than that
%of estimating species trees, because of the quantitative nature of the
%molecular data used to label the tips. In general, methods to estimate gene
%trees fall into three major categories: distance-based, maximum parsimony, and
%model-based. The first two categories are generally
%no longer used for phylogenetic inference, since model-based methods have been
%demonstrated to be almost universally more accurate. Therefore, we will review
%in detail only the last category.
%
%Model-based methods include both maximum likelihood and Bayesian approaches. To
%use them, one must first define a parametric model of sequence evolution, here
%denoted $M$. For all possible trees $T$ (with branch lengths) and choices of
%model parameters $\theta$, the model must define a probability density function
%on observed sequence data $D$, denoted $f(D \mid T,\,\theta)$. This is also
%sometimes written $\Pr(D \mid T,\,\theta)$, even though it is a probability
%\emph{density}, not a probability. Maximum likelihood aims to find a particular
%tree and values for the model parameters which optimize the \defn{likelihood
%function} $\L$ for the observed data,
%\[
%  \L(\theta,\, T \mid D) = f(D \mid T,\, \theta).
%\]
%Bayesian methods, rather than trying to find point estimates for $\theta$ and
%$T$, aim to approximate the entire \defn{posterior distribution} for the
%observed data,
%\[
%  f(T,\,\theta \mid D) = \frac{f(D \mid T,\,\theta)f(T,\,\theta)}{f(D)}.
%\]
%Here, $f(T, \theta)$ denotes the prior distribution on $T$ and $\theta$, and
%$f(D)$ is the marginal density of the data (that is, the integral of the
%numerator over all values of $T$ and $\theta$).
%
%\subsection{Estimating transmission trees}
%\label{subsubsec:treeconv}
%
%Most applications of viral phylodynamics, including this project, aim to answer
%epidemiological questions~\autocite{pybus2009evolutionary, volz2013viral}. 
%For example, phylodynamic methods have been used to investigate the degree of
%clustering~\autocite{hughes2009molecular} and the effect of elevated
%transmission risk in acute infection~\autocite{volz2012simple} 
%(a more thorough
%review is given below in subsection~\ref{subsubsec:appphylo}). For these
%applications, we are ultimately interested in the transmission tree, which is a
%representation of an epidemiological process (transmission), as opposed to the
%viral phylogeny, which represents a biological process (viral evolution). As
%discussed above in subsection~\ref{subsubsec:speciestree}, transmission trees
%and are in practice unknown, and must be estimated from available data. Here,
%we review some of the ways to perform such estimation.
%
%In principle, transmission trees can be reconstructed by on-the-ground
%epidemiological methods, that is, by asking each person who they contacted.
%This is particularly relevant for sexually transmitted diseases such as HIV,
%where individuals are more likely to recall who they contacted and
%when~\autocite{klovdahl1985social}. This kind of data is challenging to collect
%in detail. Most commonly, when applying phylodynamics in practice, the
%transmission tree is estimated from the viral
%phylogeny~\autocite{volz2013viral}. As discussed above in
%subsection~\ref{subsubsec:genetree}, the viral phylogeny is itself also estimated;
%however, for the moment we ignore this complication and describe methods for
%estimating the transmission tree if the viral phylogeny were known.
%
%When we consider the fact that the viral phylogeny is not known with certainty,
%the methods just described follow a two-step procedure. The first step is to
%estimate the viral phylogeny, and the second step is to use it to estimate the
%transmission tree.
%
%\autocite{didelot2014bayesian} develop a Bayesian version of the two-step
%approach. They first estimate a time-scaled phylogeny from viral samples, then
%estimate the transmissions which took place along the phylogeny. Due to
%within-host evolution, the transmissions are allowed to occur anywhere along
%the branches, rather than being constrained to happen at branching points.
%However, the authors admit that their method requires sampling of every
%infected individual, although they indicate that it could be extended to relax
%this assumption.
%
%A different approach is undertaken by Jombart
%\etal~\autocite{jombart2011reconstructing}, who build transmission trees
%directly from sequence data without the intermediate step of constructing a
%phylogeny. Their method finds the most likely geneology among the sampled
%sequences, assuming that the that the common ancestors to sampled isolates are
%themselves sampled. Although this may be a realistic assumption in the early
%stages of an epidemic of a slower evolving pathogen, it is unlikely for
%ancestors to be present among sequences sampled from an ongoing HIV epidemic
%\comment{Needs a reference}.
%
%Several alternative approaches have been developed which aim to
%incorporate available epidemiological data as well. Cottam
%\etal~\autocite{cottam2008integrating} first built a viral phylogeny, then
%enumerated all transmission trees consistent with this phylogeny by labelling
%the internal nodes of the phylogeny in every possible way. Each of the putative
%transmission trees was assigned a likelihood based on infection time and
%geographic separation. 
%
%

\newpage
